Abstractions for Multi-Sorted Substitutions

### 2024-03-14 paper/paper.tex
Update.

  The rest of this paper is structured as follows:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to define types
  and typing contexts for multi-sorted syntax.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservation lemmas for renamings and substitutions.
  % Section~\ref{sec:subjectreduction} introduces a small-step
  % operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:conclusion} concludes.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  % \item
  %   an adaptation of McBride's
  %   kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  %   for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which untyped substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and typing contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies in using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  \section{Types \& Contexts}
  \label{sec:types}
  % To define typing contexts and a typing relation, we need to know
  % which sorts of terms play the role of types for other sorts of terms.
  % When modeling a typed language with multi-sorted syntax, both terms
  % and types are represented as terms, but with different sorts.
  % To define typing contexts and a typing relation, we need to record for each term
  % from 
  % describe terms, and ocan be mapped to 
  % In a typed language,

@@

  \subsection{Typing Context}
  Equipped with a notion of types, we define a typing context as a
  function that maps variables to types of the same sort.

  As we want our framework to support dependent types, we allow a type returned
  by the typing context to use all variables bound previously in the typing context.
  This is achieved by the \ACode{drop-∈} function, which cuts off all
  sorts from \ACode{S} that are at smaller or equal positions than a variable \ACode{x}.
  Contexts \ACode{Γ} can be extended by a type \ACode{t} via \ACode{t \ADef{∷ᵗ} Γ}.

@@

  To use a type returned from a typing context, it needs to be
  weakened appropriately through the \ACode{\ADef{wk-telescope}} function:

@@

  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of a lambda abstraction, which
    weakens codomain type \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-kinded due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

>>>>

  The rest of this paper is structured as follows:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to define types
  and type contexts for multi-sorted syntax.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservation lemmas for renamings and substitutions.
  % Section~\ref{sec:subjectreduction} introduces a small-step
  % operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:conclusion} concludes.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  % \item
  %   an adaptation of McBride's
  %   kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  %   for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which untyped substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and type contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies in using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

@@

  \section{Types \& Contexts}
  \label{sec:types}
  % To define type contexts and a typing relation, we need to know
  % which sorts of terms play the role of types for other sorts of terms.
  % When modeling a typed language with multi-sorted syntax, both terms
  % and types are represented as terms, but with different sorts.
  % To define type contexts and a typing relation, we need to record for each term
  % from 
  % describe terms, and ocan be mapped to 
  % In a typed language,

@@

  \subsection{Type Contexts}
  Equipped with a notion of types, we define a type context as a
  function that maps variables to types of the same sort.

  As we want our framework to support dependent types, we allow a type returned
  by the type context to use all variables bound previously in the type context.
  This is achieved by the \ACode{drop-∈} function, which cuts off all
  sorts from \ACode{S} that are at smaller or equal positions than a variable \ACode{x}.
  Contexts \ACode{Γ} can be extended by a type \ACode{t} via \ACode{t \ADef{∷ᵗ} Γ}.

@@

  To use a type returned from a type context, it needs to be
  weakened appropriately through the \ACode{\ADef{wk-telescope}} function:

@@

  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor \ACode{\ACon{⊢`}}, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of a lambda abstraction, which
    weakens the codomain type \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type; and
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-kinded due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst
